---
title: Founder Effect
output: html_document
---

# (1) Sampling with replacement

Assume we have a starting frequency at 50% for "red" balls in an urn with "red" and "blue" balls.
Let's investigate what happens in the frequency of red balls with we have a large urn (600 balls) or a small urn (6 balls)

```{r}
# large urn with 600 balls, 300 red and 300 blue
nr_large <- 600; large_urn <- c( rep("red", nr_large/2), rep("blue", nr_large/2) )
```

What happens if I sample with replacement 600 balls to create the next "generation"? What is the frequency of red balls? Remember that the initial frequency is 0.50.

```{r}
length(which(sample(x=large_urn, size=nr_large, replace=TRUE)=="red")) / nr_large
```

What is the __expected__ frequency? In other words, what is the average frequency if we repeat the experiment many times? 

```{r}
freq_large <- rep(NA, 1000); for (i in 1:length(freq_large)) freq_large[i] <- length(which(sample(x=large_urn, size=nr_large, replace=TRUE)=="red")) / nr_large;
mean(freq_large)
```

Let's redo this experiment with the small urn, with only 6 balls but with still an initial frequency of 0.50 of red balls.

```{r}
# small urn with 6 balls, 3 red and 3 blue
nr_small <- 6; small_urn <- c( rep("red", nr_small/2), rep("blue", nr_small/2) )
```

What happens if we sample with replacement 6 balls to create the next "generation"? What is the frequency of red balls? Remember that the initial frequency is 0.50.

```{r}
length(which(sample(x=small_urn, size=nr_small, replace=TRUE)=="red")) / nr_small
```

What is the __expected__ frequency? In other words, what is the average frequency if we repeat the experiment many times? 

```{r}
freq_small <- rep(NA, 10000); for (i in 1:length(freq_small)) freq_small[i] <- length(which(sample(x=small_urn, size=nr_small, replace=TRUE)=="red")) / nr_small;
mean(freq_small)
```

Why?

```{r}
var(freq_large)
```

```{r}
var(freq_small)
```

# (2) Founder effect

```{r}
# ancestors (pre-colonisation)
ancestors <- rep(letters[1:20], sample(1:20,size=20,prob=20:1,replace=T))
cat("nr of ancestors:", length(ancestors))
cat("\nnr of distinct alleles:", length(unique(ancestors)))
cat("\nallele counts:"); table(ancestors)
```

```{r}
# founding event
founders <- sample(ancestors, 20)
cat("nr of founders:", length(founders))
cat("\nnr of distinct alleles:", length(unique(founders)))
cat("\nallele counts:"); table(founders)
```

```{r}
# descendants (after post-colonisation expansion)
descendants <- sample(founders, length(ancestors), replace=TRUE)
cat("nr of descendants:", length(descendants))
cat("\nnr of distinct alleles:", length(unique(descendants)))
cat("\nallele counts:"); table(descendants)
```

